---
title: 'Data for TCGAbiolinks Package'
author: Tiago C. Silva, Antonio Colaprico, Catharina Olsen, Fulvio D’Angelo, Gianluca
  Bontempi Michele Ceccarelli , and Houtan Noushmehr
date: "`r Sys.Date()`"
output:  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{'Example data for TCGA Workflow: Analyze cancer genomics and epigenomics data using Bioconductor packages'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Data Introduction

This package provides all necessary data to run and compile TCGAbiolinks package.

# Loading the data

# Data Creation

## Case study n. 3: Integration of methylation and expression for ACC

In this case of study,  we downloaded both adrenal cortical carcinoma (ACC) DNA methylation data for HumanMethylation450k platform and Gene Expression Quantification data  aligned against hg38.

Also, by default TCGAbiolinks adds subtypes already published by researchers.

We selected two molecular subtypes CIMP-low and CIMP-high to perform an integrative analysis
usng RNA expression and DNA methylation.

```{r,eval=FALSE,echo=TRUE,message=FALSE,warning=FALSE}
library(TCGAbiolinks)
library(SummarizedExperiment)
dir.create("case3")
setwd("case3")
#-----------------------------------
# STEP 1: Search, download, prepare |
#-----------------------------------
# 1.1 - DNA methylation
# ----------------------------------
query.met <- GDCquery(project = "TCGA-ACC", 
                      data.category = "DNA Methylation",
                      platform = "Illumina Human Methylation 450")
GDCdownload(query.met)
acc.met <- GDCprepare(query = query.met)

#-----------------------------------
# 1.2 - RNA expression
# ----------------------------------
query.exp <- GDCquery(project = "TCGA-ACC", 
                      data.category = "Transcriptome Profiling",
                      data.type = "Gene Expression Quantification", 
                      workflow.type = "HTSeq - Counts")
GDCdownload(query.exp)
acc.exp <- GDCprepare(query = query.exp)
```

For DNA methylation, we perform a DMR (different methylated region) analysis, 
which will give the difference of DNA methylation for the probes of the groups and their significance value.

```{r,eval=FALSE,echo=TRUE,message=FALSE,warning=FALSE}
# Volcano plot
acc.met <- TCGAanalyze_DMR(acc.met, 
                           groupCol = "subtype_MethyLevel",
                           group1 = "CIMP-high",
                           group2 = "CIMP-low",
                           p.cut = 10^-5,
                           diffmean.cut = 0.25,
                           legend = "State",
                           plot.filename = "CIMP-highvsCIMP-low_metvolcano.png")
```

For the gene expression data, we perform a DEA (differential expression analysis) which will give the fold change 
of gene expression and their significance value.

```{r,eval=FALSE,echo=TRUE,message=FALSE,warning=FALSE}
#-------------------------------------------------
# 2.3 - DEA - Expression analysis - volcano plot
# ------------------------------------------------
acc.exp.aux <- subset(acc.exp, 
                      select = colData(acc.exp)$subtype_MethyLevel %in% c("CIMP-high","CIMP-low"))

idx <- colData(acc.exp.aux)$subtype_MethyLevel %in% c("CIMP-high")
idx2 <- colData(acc.exp.aux)$subtype_MethyLevel %in% c("CIMP-low")

dataPrep <- TCGAanalyze_Preprocessing(object = acc.exp.aux, cor.cut = 0.6)

dataNorm <- TCGAanalyze_Normalization(tabDF = dataPrep,
                                      geneInfo = geneInfoHT,
                                      method = "gcContent")

dataFilt <- TCGAanalyze_Filtering(tabDF = dataNorm,
                                  qnt.cut = 0.25,
                                  method='quantile')

acc.dataDEGs <- TCGAanalyze_DEA(mat1 = dataFilt[,idx],
                            mat2 = dataFilt[,idx2],
                            Cond1type = "CIMP-high",
                            Cond2type = "CIMP-low",
                            method = "glmLRT")

save(acc.met, acc.exp, acc.dataDEGs, file = "case3.rda", compress = "xz")
```

# Session info
```{r sessionInfo, results='asis', echo=FALSE}
pander::pander(sessionInfo(), compact = FALSE)
```
